Modelling results

### 2024-09-03 notebooks/00-power-analysis-report.qmd
big progress

---
title: Power analysis report
---

@@

### Rationale

@@

### Generative model

@@

### Simulations

@@

- Number of simulations: For each effect size, we ran as many simulations as were needed to obtain a confidence interval below ±10 points around the mean. Going by steps of 50, we eventually ran 200 simulations for each effect size. These simulations ran for more than 8 hours, therefore their results are stored in `.RDS` R objects for ease of access: `data/r-data-structures/power-analyses-list.RDS` and `data/r-data-structures/power-analyses-table.RDS`. [They do not run automatically when rendering this `.qmd` file.]{.content-visible when-format="html"}

@@

### Results

@@

![Results of the power analysis using simulations. 200 simulations have been computed for each effect size.](../figures/power-analysis.png){#fig-simulations-plot}

@@

@@

cat("═════════════════════════════════════════════════════════════════════════")
cat("Session information for reproducibility:")
report_system(session = sessionInfo())
cat("Packages used:")
report_packages(session = sessionInfo())
cat("═════════════════════════════════════════════════════════════════════════")
```

>>>>

---
title: Power analysis report
subtitle: "Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm"
---

@@

# Rationale

@@

# Generative model

@@

# Simulations

@@

- Number of simulations: For each effect size, we ran as many simulations as were needed to obtain a confidence interval below ±10 points around the mean. Going by steps of 50, we eventually ran 200 simulations for each effect size. These simulations ran for more than 8 hours, therefore their results are stored in `.RDS` R objects for ease of access: `data/r-data-structures/power-analyses-list.RDS` and `data/r-data-structures/power-analyses-table.RDS`. [They do not run automatically when rendering the `.qmd` file.]{.content-visible when-format="html"}

@@

# Results

@@

![Results of the power analysis using simulations. 200 simulations have been computed for each effect size.](../figures/power-analysis.png){#suppfig-simulations-plot}

:::: {.content-visible when-format="html"}

@@

::: {.callout-note collapse="true"}
# Session information for reproducibility

@@

cat("═════════════════════════════════════════════════════════════════════════")
report_system(session = sessionInfo())
cat("Packages used:")
report_packages(session = sessionInfo())
cat("═════════════════════════════════════════════════════════════════════════")
```
:::
::::

### 2024-09-03 notebooks/01-preprocessing-report.qmd
big progress

---
title: Preprocessing report
---

```{r}
#| label: setup
#| include: false

@@

```{r}
#| label: preparing-rt-data

@@

```{r}
#| label: listing-error-rates
#| output: false

# ─── Listing error rates per subject ───
list(
  df_0e = df_0e,
  df_0i = df_0i
  ) |>
  imap(
    ~.x |>
      select(subjectid, starts_with("correct"), aphantasia) |>
      group_by(subjectid) |>
      count(pick(2)) |>
      filter(pick(1) == 1) |>
      ungroup() |>
      mutate(
        n_tot = max(n),
        prop = (n_tot - n) / n_tot * 100  # analyzing the error rate per subject
      ) %>%
      arrange(desc(prop)) |>
      select(1, 5)
    )
# 5 accuracy outliers in the explicit task, 4 in the implicit one
```

```{r}
#| label: removing-errors
#| output: false

@@

```{r}
#| label: removing-extreme-trials
#| output: false

@@

```{r}
#| label: rt-means-table
#| output: false

@@

```{r}
#| label: detecting-aberrant-means
#| output: false

@@

```{r}
#| label: removing-aberrant-means
#| output: false

@@

```{r}
#| label: preparing-correlations-data
#| output: false

@@

```{r}
#| label: filtering-extreme-groups
#| output: false

@@

@@

```{r}
#| label: session-information
#| echo: false

cat("═════════════════════════════════════════════════════════════════════════")
cat("Session information for reproducibility:")
report_system(session = sessionInfo())
cat("Packages used:")
report_packages(session = sessionInfo())
cat("═════════════════════════════════════════════════════════════════════════")
```

>>>>

---
title: Preprocessing report
subtitle: "Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm"
eval: false
---

```{r}
#| label: setup
#| include: false
#| eval: true

@@

```{r}
#| label: preparing-rt-data
#| code-summary: Preparing the data for the outlier detection procedure

@@

```{r}
#| label: listing-error-rates
#| code-summary: Listing error rates per subject

# ─── Listing error rates per subject ───
list(
  df_0e = df_0e,
  df_0i = df_0i
  ) |>
  imap(
    ~.x |>
      select(subjectid, starts_with("correct"), aphantasia) |>
      group_by(subjectid) |>
      count(pick(2)) |>
      filter(pick(1) == 1) |>
      ungroup() |>
      mutate(
        n_tot = max(n),
        prop = (n_tot - n) / n_tot * 100  # analysing the error rate per subject
      ) %>%
      arrange(desc(prop)) |>
      select(1, 5)
    )
# 5 accuracy outliers in the explicit task, 4 in the implicit one
```

```{r}
#| label: removing-errors
#| code-summary: Removing error trials and irrelevant variables

@@

```{r}
#| label: removing-extreme-trials
#| code-summary: Removing extreme RTs

@@

```{r}
#| label: rt-means-table
#| code-summary: Creating a table with RT means

@@

```{r}
#| label: detecting-aberrant-means
#| code-summary: Detecting outliers in the RT means

@@

```{r}
#| label: removing-aberrant-means
#| code-summary: Removing outliers in the RT means

@@

```{r}
#| label: preparing-correlations-data
#| code-summary: Preparing the data for the correlation analysis

@@

```{r}
#| label: filtering-extreme-groups
#| code-summary: Creating extreme groups

@@

:::: {.content-visible when-format="html"}

@@

::: {.callout-note collapse="true"}
# Session information for reproducibility

```{r}
#| label: session-information
#| echo: false
#| eval: true

cat("═════════════════════════════════════════════════════════════════════════")
report_system(session = sessionInfo())
cat("Packages used:")
report_packages(session = sessionInfo())
cat("═════════════════════════════════════════════════════════════════════════")
```
:::
::::

### 2024-09-03 notebooks/02-modelling-results.qmd
big progress

---
title: Modelling results
---

@@

library(here)
source(here("scripts/brms-models.R"))
```

@@

# Implicit task figures

```{r}
summary(fit_8)
```

```{r}
#| label: estimated-means-fit-8

emmeans_8 <- 
  emmeans(fit_8, c("aphantasia", "congruence"), epred = TRUE) |> 
  summary(point.est = median) |> 
  as_tibble() |> 
  mutate(across(3:5, ~.*1000)) |>
  rename(median = emmean)
```

```{r}
subj_means_8 <-
  df_i_rt |> 
  group_by(subjectid, aphantasia, congruence) |>
  reframe(rt = median(rt)*1000)
```

Without individual subject lines:

```{r}
#| label: fig-8
#| fig-width: 9
#| fig-height: 6

dw <- .4

subj_means_8 |> 
  ggplot(aes(x = congruence, y = rt, color = aphantasia, fill = aphantasia)) +
  # ggplot(aes(x = aphantasia, y = rt, color = congruence, fill = congruence)) +
  geom_violinhalf(
    trim = TRUE,
    flip = c(1, 2),
    alpha = .5,
    scale = "width",
    position = position_dodge(width = dw),
    show.legend = FALSE
  ) +
  geom_quasirandom(
    color = "white",
    pch = 21,
    width = .05,
    size = 1.8,
    alpha = .6,
    show.legend = FALSE,
    dodge.width = .4
  ) +
  # group median lines
  geom_line(
    data = emmeans_8,
    aes(y = median, group = aphantasia),
    position = position_dodge(width = dw),
    size = 1,
    show.legend = FALSE
  ) +
  geom_errorbar(
    data = emmeans_8,
    aes(y = median, ymin = lower.HPD, ymax = upper.HPD),
    position = position_dodge(width = dw),
    width = 0,
    linewidth = 1,
    color = "black",
    show.legend = FALSE
  ) +
  geom_point2(
    data = emmeans_8,
    aes(y = median),
    position = position_dodge(width = dw),
    size = 3,
    shape = 21,
    stroke = 1,
    color = "black"
  ) +
  labs(x = NULL, fill = NULL) +
  scale_y_continuous(
    name = "Response time (ms)",
    breaks = breaks_pretty()
  ) +
  theme(
    axis.ticks.y = element_line(linewidth = 0.5),
    panel.grid.major.y = element_line(linewidth = 0.5, color = "grey90"),
    panel.grid.minor.y = element_line(linewidth = 0.5, color = "grey95"),
    legend.position = "top"
  )
```

With individual subject lines:

```{r}
#| label: fig-8-with
#| fig-width: 10
#| fig-height: 6

alpha_subj <- .3
nudge <- .11

subj_means_8 |> 
  ggplot(aes(x = congruence, y = rt, color = aphantasia, fill = aphantasia)) +
  # ggplot(aes(x = aphantasia, y = rt, color = congruence, fill = congruence)) +
  geom_violinhalf(
    trim = TRUE,
    flip = c(1, 2),
    alpha = .5,
    scale = "width",
    position = position_dodge(width = dw),
    show.legend = FALSE
  ) +
  geom_quasirandom(
    color = "white",
    pch = 21,
    width = .05,
    size = 1.8,
    alpha = .6,
    show.legend = FALSE,
    dodge.width = .4
  ) +
  # individual subject lines
  geom_line(
    data = subj_means_8 |> filter(aphantasia == "Aphantasia"),
    aes(group = interaction(subjectid,aphantasia)),
    position = position_nudge(nudge),
    alpha = alpha_subj,
    show.legend = FALSE
  ) +
    geom_line(
    data = subj_means_8 |> filter(aphantasia == "Control"),
    aes(group = interaction(subjectid, aphantasia)),
    position = position_nudge(-nudge),
    alpha = alpha_subj,
    show.legend = FALSE
  ) +
  # # group median lines
  # geom_line(
  #   data = emmeans_8,
  #   aes(y = median, group = aphantasia),
  #   position = position_dodge(width = dw),
  #   size = 1,
  #   show.legend = FALSE
  # ) +
  geom_errorbar(
    data = emmeans_8,
    aes(y = median, ymin = lower.HPD, ymax = upper.HPD),
    position = position_dodge(width = dw),
    width = 0,
    linewidth = 1,
    color = "black",
    show.legend = FALSE
  ) +
  geom_point2(
    data = emmeans_8,
    aes(y = median),
    position = position_dodge(width = dw),
    size = 3,
    shape = 21,
    stroke = 1,
    color = "black"
  ) +
  labs(x = NULL, fill = NULL) +
  scale_y_continuous(
    name = "Response time (ms)",
    breaks = breaks_pretty()
  ) +
  theme(
    axis.ticks.y = element_line(linewidth = 0.5),
    panel.grid.major.y = element_line(linewidth = 0.5, color = "grey90"),
    panel.grid.minor.y = element_line(linewidth = 0.5, color = "grey95"),
    legend.position = "top"
  )
```

```{r}
#| label: fig-8-zoom
#| fig-width: 10
#| fig-height: 6

subj_means_8 |> 
  ggplot(aes(x = congruence, y = rt, color = aphantasia, fill = aphantasia)) +
  geom_line(
    data = emmeans_8,
    aes(y = median, group = aphantasia),
    position = position_dodge(width = .5),
    size = 1.5,
    show.legend = FALSE
  ) +
  geom_errorbar(
    data = emmeans_8,
    aes(y = median, ymin = lower.HPD, ymax = upper.HPD),
    position = position_dodge(width = .5),
    width = 0,
    linewidth = 1,
    color = "black",
    show.legend = FALSE
  ) +
  geom_point2(
    data = emmeans_8,
    aes(y = median),
    position = position_dodge(width = .5),
    size = 5,
    shape = 21,
    stroke = 1,
    color = "black"
  ) +
  labs(x = NULL, fill = NULL) +
  scale_y_continuous(
    name = "Response time (ms)",
    breaks = breaks_pretty(n = 8)
  ) +
  theme(
    axis.text.y = element_text(size = 10),
    # axis.ticks.y = element_line(linewidth = 0.5),
    panel.grid.major.y = element_line(linewidth = 0.5, color = "grey90"),
    panel.grid.minor.y = element_line(linewidth = 0.5, color = "grey95"),
    legend.position = "top"
  )
```

# Explicit task figures

>>>>

---
title: Modelling results
subtitle: "Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm"
---

@@

library(here)
source(here("scripts/00-preprocessing.R"))
```

@@

# Questionnaires analyses

# Accuracy analyses

# Response Times analyses

## Rationale

To account for the non-normal, positively skewed distributions of the RTs, we fitted Generalized Linear Mixed Models (GLMMs) with inverse Gaussian distributions. The models were implemented in the `lme4` R package and integrated in `tidymodels` workflows using the package `multilevelmod`. Models with Gamma and Gaussian distributions were also fitted and compared with the AIC and BIC to ensure that we chose the best distribution available.

The models included the ***Group*** (aphantasic, control), ***Congruence*** condition (congruent or incongruent) and ***Color*** condition (color or uncolored) along with all their two and three way interactions as fixed categorical predictors, while ***participants*** have been included as grouping factors (i.e. "random effects"). The random effect structure was chosen by fitting and comparing models with every possible combination of distribution and structure (intercept by participant, congruence or color, slope by participant on congruence and/or color) aiming for the best balance between goodness of fit and parsimony. Complex random-effects structures including various slopes on the factors failed to converge to stable and reliable estimates, hence the optimal models chosen included a single by-participant random intercept.

## Model fitting

The formula of the model fitted is $RT \sim (aphantasia + congruence + color)^3 + (1|subjectid)$. [See the HTML version for the code and details.]{.content-hidden when-format="html"}

::: {.content-visible when-format="html"}
```{r}
#| label: model-fitting
#| code-fold: false

model_formula <- rt ~ (aphantasia + congruence + color)^3 + (1|subjectid)

fit_implicit <- 
  glmer(
    formula = model_formula,
    data = df_i_rt,
    family = inverse.gaussian(link = "identity"),
    control = glmerControl(optimizer = "bobyqa")
  )

fit_explicit <- 
  glmer(
    formula = model_formula,
    data = df_e_rt,
    family = inverse.gaussian(link = "identity"),
    control = glmerControl(optimizer = "bobyqa")
  )
```
:::

## Model diagnostics

The quality checks of the models are displayed in [@suppfig-implicit-checks and @suppfig-explicit-checks]{.content-visible when-format="html"} [@suppfig-implicit-checks-2 and @suppfig-explicit-checks-2]{.content-hidden when-format="html"}.

:::: {.content-visible when-format="html"}
::: {.panel-tabset .column-page-inset}

### Implicit task

```{r}
#| label: implicit-checks
#| fig-width: 14
#| fig-height: 12
#| eval: false

# characteristics to check
model_checks <- c("pp_check","homogeneity", "vif", "outliers", "qq", "reqq")

check_model(
  fit_implicit, 
  checks = model_checks, 
  detrend = FALSE, 
  residual_type = "normal"
  ) |> 
  plot()

ggsave(here("figures/checks-implicit.png"), dpi = 600)
```

![Model assumption checks for the Generalized Linear Mixed Model fit on the RTs in the implicit task.](../figures/checks-implicit.png){#suppfig-implicit-checks}

### Explicit task

```{r}
#| label: explicit-checks
#| fig-width: 14
#| fig-height: 12
#| eval: false

check_model(
  fit_explicit, 
  checks = model_checks, 
  detrend = FALSE, 
  residual_type = "normal"
  ) |> 
  plot()

ggsave(here("figures/checks-explicit.png"), dpi = 600)
```

![Model assumption checks for the Generalized Linear Mixed Model fit on the RTs in the explicit task.](../figures/checks-explicit.png){#suppfig-explicit-checks}

:::
::::

:::{.content-hidden when-format="html"}

![Model assumption checks for the Generalized Linear Mixed Model fit on the RTs in the implicit task.](../figures/checks-implicit.png){#suppfig-implicit-checks-2}

![Model assumption checks for the Generalized Linear Mixed Model fit on the RTs in the explicit task.](../figures/checks-explicit.png){#suppfig-explicit-checks-2}

:::

## Model summaries

:::{#supptbl-glmm-i-summary}
```{r}
#| label: summary-implicit
#| echo: false

model_performance(fit_implicit) |> display()

model_parameters(fit_implicit) |> display()
```

Performance and estimates of the GLMM fitted on the implicit task data.
:::

:::{#supptbl-glmm-e-summary}
```{r}
#| label: summary-explicit
#| echo: false

model_performance(fit_explicit) |> display()

model_parameters(fit_explicit) |> display()
```

Performance and estimates of the GLMM fitted on the explicit task data.
:::

## Estimated means and contrasts

:::{#supptbl-estimates-implicit}
```{r}
#| label: emmeans-implicit
#| echo: false

emmeans_im <-
  emmeans(fit_implicit, c("aphantasia", "congruence")) |>
  summary(point.est = median) |>
  as_tibble() |>
  mutate(across(3:7, ~.*1000)) |>
  rename(median = emmean)

emmeans_im |> 
  rename(Group = 1, Condition = 2, "Median (ms)" = median) |>
  display()

fit_implicit |> 
  estimate_contrasts(
    contrast = c("aphantasia", "congruence"), 
    p_adjust = "none"
    ) |> 
  display()
```

Estimated means for each group in each congruence condition and contrasts between groups and conditions in the implicit task.
:::

:::{#supptbl-estimates-explicit}
```{r}
#| label: emmeans-explicit
#| echo: false

emmeans_ex <-
  emmeans(fit_explicit, c("aphantasia", "congruence")) |>
  summary(point.est = median) |>
  as_tibble() |>
  mutate(across(3:7, ~.*1000)) |>
  rename(median = emmean)

emmeans_ex |> 
  rename(Group = 1, Condition = 2, "Median (ms)" = median) |>
  display()

fit_explicit |> 
  estimate_contrasts(
    contrast = c("aphantasia", "congruence"), 
    p_adjust = "none"
    ) |> 
  display()
```

Estimated means for each group in each congruence condition and contrasts between groups and conditions in the explicit task.
:::

## Visualisations

The figures below are also displayed in the main article. [See the HTML version for the code and details.] {.content-hidden when-format="html"}

```{r}
#| label: fig-im
#| fig-cap: Subject means and model-estimated means per group and condition in the implicit task.
#| fig-width: 18
#| fig-height: 7

subj_means_im <-
  df_i_rt |> 
  group_by(subjectid, aphantasia, congruence) |>
  reframe(rt = mean(rt)*1000)

(
  plot_models_full(subj_means_im, emmeans_im, y_min = 400, y_max = 1110, size = 4) +
  plot_models_zoomed(subj_means_im, emmeans_im, size = 4) &
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 20),
    legend.margin = margin(0, 150, 25, 0),
    axis.title.y = element_text(size = 20),
    axis.text.x = element_text(size = 18)
  )
) +
  plot_layout(
    guides = "collect",
    widths = c(1, 1.2)
  )

# ggsave(here("figures/model-implicit.png"), dpi = 600)
```

```{r}
#| label: fig-ex
#| fig-cap: Subject means and model-estimated means per group and condition in the explicit task.
#| fig-width: 18
#| fig-height: 7

subj_means_ex <-
  df_e_rt |> 
  group_by(subjectid, aphantasia, congruence) |>
  reframe(rt = mean(rt)*1000)

(
  plot_models_full(subj_means_ex, emmeans_ex, y_min = 350, y_max = 1450, size = 4) +
  plot_models_zoomed(subj_means_ex, emmeans_ex, size = 4) &
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 20),
    legend.margin = margin(0, 150, 25, 0),
    axis.title.y = element_text(size = 20),
    axis.text.x = element_text(size = 18)
  )
) +
  plot_layout(
    guides = "collect",
    widths = c(1, 1.2)
  )

# ggsave(here("figures/model-explicit.png"), dpi = 600)
```

# Correlational analyses

:::: {.content-visible when-format="html"}
